Mathematical Aspects of the Capital Market

### 2014-06-11 thesis/thesis.tex
Move matlab files to a separate folder.

\documentclass{report}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{amssymb}
\usepackage{amsfonts, amsmath}
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}

\input{../physics_common.tex}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% \title{Stocks' Cross-Correlations}
% \subtitle{}

\author{Xie Xiaolei}
\date{\today}
\begin{document}

@@

% Title
%\HRule \\[0.4cm]
{ \huge \bfseries Returns Models and Their Covariance Matrix
  \\[0.4cm] }

@@

\end{center}
\end{titlepage}
% \maketitle
\tableofcontents

@@

Here by dynamics we mean how the value of an asset is affected by its
own history as well as by the histories and current values of other
assets in the market. The influence from the asset's own history is
termed autocorrelations, i.e. correlations in time, while
the influence from other assets are termed cross correlations.

Mathematical models that describe the aforementioned observables in
terms of probability distributions, autocorrelations and cross
correlations give predictive power and help determine a fair price of
a given asset. Hence they are a central topic of the mathematics of
the financial market. In the domain of discrete-time returns' models,
GARCH and stochastic volatility models (SV) are the most popular, so in
this thesis project we study a total of 7 intraday return series 
using GARCH and SV models and compare their performance in volatility
forecast.

% Once a model has been tentatively constructed, it must be
% verified. The verification again relies on the statistics of the
% observables. Over the years, a number of phenomena have been
% consistently reported and become generally accepted. These are referred
% to as stylized facts, which are detailed in \S\ref{sec:StylizedFacts}
% and constitute a standardized test suit for model verification.

% Judged by the stylized facts, each of the models proposed in the
% literature has its own advantages and disadvantages. The most
% prominent and influential among them are arguably the GARCH
% models. Their first and simplest variant was proposed by Bollerslev in
% 1990 \cite{Bollerslev86} and has been developed tremendously since
% then.

% An apparent advantage of the GARCH models is their simplicity. They
% are linear models, where the conditional variance is modeled as a
% linear combination of past variances and past innovations squared. As
% a result, it is relatively easy to fit a model to a given set of data,
% i.e. to estimate the model parameters with respect to the
% data. Because of the linearity, the estimation of model parameters
% often give consistent results, regardless of the choice of initial
% parameter values or the choice of pre-sample data.

% In contrast, many of the other types of models are often specified in
% a complicated, although more advanced way, for example, by stochastic
% differential equations. This kind of specifications certainly have the
% advantage of directly linking to the theory of stochastic processes,
% but nonetheless are difficult to fit to data, one of the most tricky
% problems being the high sensitivity to the choice of initial parameter
% values.

% Another advantage of GARCH models is that their tail behavior is
% better understood than that of their competitors. The simplest variant
% of GARCH, namely the GARCH(1,1) models are proven to have regularly
% varying tails, or in common terms, power-law tails
% \cite{mikosch2000}. In particular, if the $\alpha_1$ and $\beta_1$
% parameters of the model sum up to 1, the return series generated by
% the model has a power-law tail with power exponent 2, and consequently
% the squared returns have a tail with power exponent 1 -- by the
% central limit theorem the sum of the squared returns will converge to
% the Cauchy distribution, a stable distribution with L\'evy index 1. We
% explore this fact in more details in chapter
% \ref{chp:CrossCorrelationFat} where the cross-correlation matrix is
% studied.

% The GARCH model with $\alpha_1 + \beta_1 = 1$ is more often referred
% to as an IGARCH model, aka. an integrated GARCH model. It is of
% particular interest because, very often, when fitting a GARCH(1,1)
% model to real financial data, one finds $\alpha_1 + \beta_1$ extremely
% close to 1, i.e. at values like 0.99, 0.995, etc.

% A third advantage of GARCH models is their flexibility. Namely one can
% choose any appropriate distribution to model the innovations. While
% the Gaussian and the student's t distributions are the most common,
% the use of other distributions such as asymmetric student's t and
% Johnson Su has also been studied in the literature (see
% \cite{Simonato2012}).

% However, GARCH models have their disadvantages too. The original
% GARCH model does not account for skewness in the distribution of
% the returns. As a result, the skewness, often attributed to
% price-volatility correlations, have to be added ``by hand''. This kind
% of treatment fixes the model on the one hand, but introduces some
% arbitrariness on the other. Put in another way, skewness does not
% follow naturally from the structure of the model but rather is added
% explicitly.

% Moreover, fitting a GARCH model often results in a large number of
% parameters; 16 to 30 are not uncommon when intra-day data are being
% dealt with. This raises the question of true fitness: the model
% specification may indeed fit the selected data set but very well fall
% short of an accurate description of the time series in question. In
% other words, such a model is unstable and hence unreliable: if one
% removes a small fraction of data from the beginning of the set and
% appends some to the end -- a necessary operation for the evolution of
% the model -- the model parameters change drastically and so do the
% predictions.

% As intra-day transaction data become more accessible, the volatility,
% which has been considered an unobservable quantity by models like
% GARCH, can be estimated by the so called realized volatility, which is
% the square root of the sum of squared returns that have been sampled
% at a higher frequency than are those that one intends to model. For
% example, it has been shown that the volatility of daily returns is
% best estimated using 17-minute intra-day returns \cite{Sahalia05}.

% Now that the volatility can be estimated, models that take advantage
% of this are expected to achieve significantly higher accuracy than
% those that do not. The model presented in this thesis is built indeed
% in this spirit.

% Like the GARCH models, our proposed model has the simple structure
% $r_t = \sigma_t z_t$, where $z_t$ has unit variance and is
% drawn from any appropriate distribution; $\sigma_t$ is the
% volatility. Unlike GARCH, which treats $\sigma_t$ as deterministic at
% time $t$, we treat $\sigma_t$ as random and model $\ln\sigma_t$ using
% time series analysis. This is possible because past values of
% $\ln\sigma_t$ can be estimated using past values of the returns at a
% higher frequency, as has just been discussed.

% Our model has a number of advantages too. Like the original
% GARCH model, it is simple. It has two random variables $z_t$ and
% $\ln\sigma_t$, the latter being the subject of time series
% modeling. The use of conventional time series analysis means the
% model can be not only more accurate but also more economic -- long
% memory of the volatility, which is the cause of a large number of
% parameters in a GARCH model, can be modeled by differencing or
% fractional differencing. As a result, much fewer parameters are needed
% for the model.

% Secondly, unlike GARCH models, skewness follows directly from the
% correlation between the two random variables $z_t$ and
% $\ln\sigma_t$. As shown in chapter \ref{chp:PriceModels},
% positive/negative correlation leads to positive/negative skewness,
% and skewness increases with the strength of correlation.

% Thirdly, our model has sub-exponential tails, as shown in chapter
% \ref{chp:PriceModels}. This type of tails are not as fat as regularly
% varying tails, but in our specific case, the tails do behave
% approximately as a power-law in the regions relevant to stock and
% index returns. After all, power-law behavior is an empirical result
% obtained in the returns data, and such data are not available to a
% statistically significant amount at extremely large values. In plain
% words, situations where one makes a 20\%+ profit on a single stock
% trade in a single day happen too rarely to allow any statistical
% statement.

Another important part of the thesis is about the cross-correlation
matrix, for which we study the distribution of its elements as well
as its eigenvalues. When the cross-correlation matrix is constructed
from returns with simple Gaussian distribution, the matrix is called a
Wishart Matrix and has been studied extensively in the literature. If
the returns have L\'evy distributions, the matrix is termed
Wishart-L\'evy and has been studied to some extent, particularly
regarding its eigenvalue distribution \cite{politi2010}.

However, it is understood that real stock/index returns are much more
complicated than a straight-forward Gaussian or L\'evy distribution can
describe --- instead, one needs structured models for their description.

Therefore in this thesis we study how the elements and the eigenvalues
of the cross-correlation matrix are distributed when the matrix is
built from returns described by different returns' models, specifically
GARCH(1,1) and stochastic volatility models, whose the tail behavior is
understood. Moreover, we also study how auto-correlations in the
returns influence the aforementioned distributions --- such
auto-correlations, known as second-order auto-correlations decay
exponentially but may still leave footprints in the cross-correlation
matrix.

The rest of this report is organized as follows:

% \S\ref{sec:FundamentalConcepts} presents a few very important
% concepts and notations that will often be referred to in later
% chapters; \S\ref{sec:StylizedFacts} gives a review of the ``stylized
% facts'', i.e. a number of consistently reported and generally accepted
% phenomena that must be accounted for by a returns' model.

Chapter \ref{chp:PriceModels} reviews some of the most
influential returns' models, fits them to a few intraday return
series and compares their performance. A calculation of the
unconditional distribution functions of stochastic volatility models,
especially in the case where the residual of the log-volatility
and the innovation of the return are correlated normal variates, is
also presented. Chapter \ref{chp:Gaussian} investigates the elements' and
eigenvalues' distributions of the Wishart matrix, and studies the influence
of auto-correlated returns. Chapter \ref{chp:CrossCorrelationFat}
looks at the elements' and eigenvalues' distributions of a covariance
matrix that is built from identically specified GARCH(1,1) series.

@@

\chapter{Covariance Matrix of GARCH(1,1) Returns}
\label{chp:CrossCorrelationFat}

% In this chapter we study the cross-correlation matrix $C$ constructed from
% returns that have a fat-tailed distribution, i.e.
% \begin{eqnarray*}
%   C &=& {1 \over T^{2/\alpha}} RR'
% \end{eqnarray*}
% where $R$ is an $N\times T$ returns' matrix whose entries are
% \begin{eqnarray}\label{eq:CrossCorrelationMatrix}
%   R_{it} &=& \left\{
%   \begin{array}{ll}
%     a_{it} & \text{if $t=1$} \\
%     \phi R_{i, t-1} + a_{it} & \text{if $t>1$} \\
%   \end{array} \right.
% \end{eqnarray}
% Here $a_{it}$ are random variables drawn from a stable distribution
% with l\'evy index $\alpha$. In \S\ref{sec:LevyWishart} we look at
% the case where the stable distribution is chosen to be Cauchy; in
% \S\ref{sec:XieWishart} we look at the case where $a_{it}$ is
% described by an ARIMA log-volatility model, which is detailed in
% \S\ref{sec:SLV_model}; finally in \S\ref{sec:GarchWishart} we
% investigate the case where $a_{it}$ is a GARCH(1, 1) process, which is
% proven to have regularly varying tails by Mikosch et al
% \cite{mikosch2000}.

% \section{Wishart-L\'evy Matrices}\label{sec:LevyWishart}
% In this section we study the case where the distribution of $a_{it}$
% as appear in equation \ref{eq:CrossCorrelationMatrix} is chosen to be
% Cauchy, a stable distribution with L\'evy index $\alpha =
% 1$. Politi et al derived the eigenvalue distribution of Wishart-L\'evy
% matrices in \cite{politi2010} using free probability theory. Here we
% show, qualitatively, how the distribution changes when
% auto-correlations are present among the return series. To be precise,
% the PDF of $a_{it}$ is
% \begin{eqnarray*}
%   f_a(x) &=& {\gamma \over \pi} {1 \over \gamma^2 + x^2}
% \end{eqnarray*}

% By numerically simulating a large number of cross-correlation matrices
% and solving for the eigenvalues of each simulated matrix, one can
% obtain the eigenvalue distribution corresponding to each value of
% $\phi$. Figure \ref{fig:WishartLevySpectraPositive} shows the
% distribution for a range of positive $\phi$ values, and figure
% \ref{fig:WishartLevySpectraNegative} shows those for a range of
% negative $\phi$ values.
% \begin{figure}[htb!]
%   \centering
%   \subfigure[]{
%     \includegraphics[scale=0.35, clip=true, trim=94 223 109
%     134]{../pics/WishartLevyEigDist.pdf}
%     \label{fig:WishartLevySpectraPositive}
%   }
%   \subfigure[]{
%     \includegraphics[scale=0.35, clip=true, trim=94 223 109
%     134]{../pics/WishartLevyNegEigDist.pdf}
%     \label{fig:WishartLevySpectraNegative}
%   }
%   \caption{\small \it Eigenvalue distribution (PDF) of Wishart-L\'evy
%     Matrices. Horizontal axis: eigenvalues ($\lambda$) on log-scale;
%     Vertical axis: PDF ($f_\phi(\lambda)$) of the eigenvalues on
%     log-scale. \ref{fig:WishartLevySpectraPositive}: $\phi > 0$;
%     \ref{fig:WishartLevySpectraNegative}: $\phi < 0$;  The $|\phi|$
%     values 0, 0.5, 0.8, and 0.99 correspond to $\tau$ (correlation
%     time) values 0, 1.0, 3.10, and 68.97.  The correlation time is
%     independent of the sign of $\phi$.}
% \end{figure}

% It is seen in these figures
% \begin{itemize}
% \item the eigenvalue distribution is identical for $\phi$ and
%   $-\phi$;
% \item the distribution is not bounded within a finite interval but rather
%   extends to infinity, in constrast to the Marcenko-Pastur law that
%   bounds the maximum eigenvalue at $\sigma^2(1 + \sqrt{N/T})^2$;
% \item as auto-correlation increases, the distribution flatens,
%   implying probability being transferred to large eigenvalues.
% \end{itemize}

% \section{Cross-Correlation Matrix of ARIMA Log-volatility
%   returns}\label{sec:XieWishart}
% In this section we study the cross-correlation matrix of return series
% specified as follows:
% \begin{equation}
%   \label{eq:CrossCorrSLV1}
%   \begin{aligned}
%     R_{it} &= \phi R_{i, t-1} + a_{it} \\
%     a_{it} &= e^{v_{it}} b_{it}
%   \end{aligned}
% \end{equation}
% where $v_{it}$ is typically described by an ARIMA model as discussed
% in \S\ref{sec:SLV_model}. The unconditional distribution of
% such returns is calculated in \S\ref{sec:XieCalc}, where it is shown
% that all the moments of the distribution are finite. Based on this
% observation, we expect the Marcenko-Pastur law to emerge for the
% eigenvalue distribution of the cross-correlation matrix C, when
% $\phi=0$, i.e. when no auto-correlations are absent. In this case C is
% built as
% \begin{eqnarray*}
%   C &=& {1 \over T} \sum_{t=1}^T (DR)(DR)' \\
% \end{eqnarray*}
% where $R'$ denotes the transpose of $R$ and
% \begin{eqnarray*}
%   D &=&
%   \begin{pmatrix}
%     1/\sigma_1 & 0 & \cdots & 0 \\
%     0 & 1/\sigma_2 & \cdots & 0 \\ 
%     \vdots & \vdots & \ddots & 0 \\
%     0 & 0 & \cdots & 1/\sigma_N
%   \end{pmatrix}
% \end{eqnarray*}
% $\sigma_i$ is the standard deviation of return series $i$. By
% numerical simulation as described above, the eigenvalue distribution
% of C is obtained. We plot it in figure \ref{fig:FatWishartSpectra}.
% \begin{figure}[htb!]
%   \centering
%   \vspace{-20mm}
%   \includegraphics[scale=0.5, clip=true, trim=91 229 116
%   137]{../pics/FatWishartEigDist.pdf}
%   \caption{\small \it Eigenvalue distribution (PDF) of a
%     cross-correlation matrix with returns described by
%     \ref{eq:CrossCorrSLV1}. $\phi$ = 0, 0.5, 0.8, 0.95 correspond to
%     correlation time $\tau$ = 0.00, 1.00, 3.11, 13.51}
%   \label{fig:FatWishartSpectra}
% \end{figure}
% As expected, the case with zero auto-correlation has an eigenvalue
% distribution governed by the Marcenko-Pastur law. As auto-correlations
% become stronger, the distribution deforms, but in a rather different
% way if compared to the case of Gaussian returns. Figure
% \ref{fig:FatWishartSpectra} shows that the minimum eigenvalue
% decreases with increasing $\phi$ in the present case of ARIMA
% log-volatility returns, while figure \ref{fig:Gaussian_mineig} shows
% that, in the case of Gaussian returns, the minimum eigenvalue actually
% increases with $\phi$. We leave this interesting comparison to future
% investigations.

% \section{Cross-Correlation Matrix of GARCH(1,1) returns}
% \label{sec:GarchWishart}
GARCH models and particularly GARCH(1,1) models are widely used to
model financial return series of various time scales, ranging from
daily and monthly returns that have been studied extensively in the
literature to intraday returns that we have selectively investigated
in chapter \ref{chp:PriceModels}. One particularly nice feature of
GARCH(1,1) models is that they have regularly varying tails (power-law
tails) \cite{Mikosch2009, mikosch2000} even when the innovations
(denoted $z_t$ in the following text) are normally distributed ---
something not shared by other classes of models, e.g. not by ARIMA
log-volatility models, as shown in \S\ref{sec:XieCalc}. However, what
this tail behavior implies for the covariance matrix is much less
understood, especially when the constituent return series are
auto-correlated.

So in this chapter we consider the covariance matrix of N
identically specified, possibly auto-correlated GARCH(1, 1)
processes:
\begin{eqnarray}
  r_{it} &=& \phi r_{i, t-1} + \epsilon_{it} \nonumber \\
  \epsilon_{it} &=& \sigma_{it} z_{it} \label{eq:garch_spec}
\end{eqnarray}
where $i=1,2,...,N$; $t=1,2,...,T$; $z_{it}$ is independent,
identically distributed, and
\begin{eqnarray*}
  \sigma_{it}^2 &=& \alpha_0 + \alpha_1 z_{i, t-1}^2 + \beta_1
  \sigma_{i,t-1}^2
\end{eqnarray*}
Mikosch et al showed in \cite{mikosch2000} that a GARCH(1,1) process
satisfying
\begin{eqnarray*}
  \alpha_0 &>& 0 \\
  \E(\alpha_1 Z^2 + \beta_1) &<& 0 \\
\end{eqnarray*}
and
\begin{eqnarray*}
  \E[(\alpha_1 z^2 + \beta_1)^{p/2}] &\geq& 1 \\
  \E|Z|^p \ln|Z| &\leq& \infty
\end{eqnarray*}
for some $p > 0$, is stationary and has regularly varying tails. The
tail exponent $\alpha$ is determined by:
\begin{equation}\label{eq:garch_alpha}
  \E[(\alpha_1 Z^2 + \beta_1)^{\alpha/2}] = 1
\end{equation}
Here $Z$ is a random variable that has the same distribution as
$z_{it}$. In our simulations described hereafter $z_{it}, Z \sim N(0,
1)$. In this section and the next, we first study situations where no
auto-correlations are present among the returns, i.e. $\phi = 0$; then
in \S\ref{sec:garch_nonzero_phi} we look at how auto-correlations
change the picture.

>>>>

\documentclass{report}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{amssymb}
% \usepackage{amsfonts, amsmath}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage[bookmarks=true]{hyperref}
\usepackage[acronym]{glossaries}
\usepackage{bookmark}

\input{../physics_common.tex}
\input{./acronyms.tex}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% \title{Stocks' Cross-Correlations}
% \subtitle{}
\makeglossaries
\author{Xie Xiaolei}
\date{\today}
\begin{document}

@@

% Title
%\HRule \\[0.4cm]
{ \huge \bfseries Return Models and Covariance Matrices
  \\[0.4cm] }

@@

\end{center}
\end{titlepage}
% \maketitle
\begin{abstract}
Return models and covariance matrices of return series have been
studied. In particular, \gls{garch} and \gls{sv} models
are compared with respect to their forecasting accuracy when applied
to intraday return series. \gls{sv} models are found to be considerably more
accurate and more consistent in accuracy in forecasting.

Covariance matrices formed from Gaussian and \gls{garch} return series, and
in particular, return series auto-correlated as an AR(1) process, have
been studied. In the case of Gaussian returns, the largest eigenvalue
is found to approximately follow a gamma distribution also when the
returns are auto-correlated. Expressions relating the mean and the
variance of the asymptotic Gaussian distribution of the matrix
elements are derived. In the case of \gls{garch} returns, both the largest
and the smallest eigenvalues of the covariance matrix are seen to
increase with increasing auto-correlation. The matrix elements are
found to follow L\'evy distributions with different L\'evy indices for
the diagonal and the non-diagonal elements.

Localization of eigenvectors of covariance matrices of
returns from \gls{garch} processes has been investigated. It is
found that the localization is reduced as the auto-correlation is
increased. Quantitatively, the number of localized eigenvectors
decreases approximately as a quadratic function with the
auto-correlation strength, i.e. the autoregressive coefficient of the
AR(1) process.
\end{abstract}

\section*{Acknowledgements}
First of all, I thank my supervisor, Prof. Sven \AA berg, for the many
hours of mentoring, discussion, encouragement and guidance. Without
them, this report will not exist. Then I wish to give my gratitude to
the people in the division of mathematical physics. The casual
discussions that I had with them truely inspired me, and the friendly
and lively atmosphere that they create has helped bring peace and clarity
to my mind.

Last but not least, I want to thank my wife, Wang Xiaodan, for her
unconditional support and understanding.

\tableofcontents

% \chapter*{List of Acronyms}

\printglossaries

@@

The dynamics of an asset is affected by its own history as well as by
the histories and current values of other assets in the market. The
influence from the asset's own history is termed autocorrelations,
i.e. correlations in time, while the influence from other assets are
termed cross correlations.

\Gls{sv} models built on historical data describe the time evolution
of the aforementioned observables in terms of probability
distributions, autocorrelations and cross correlations. They may have
some predictive power and help determine a fair price of a given
asset.

Quite often, instead of an asset's price, the relative price change,
i.e. the return, is studied. Analytically solvable models often assume
Gaussian return distribution, although data suggest fatter tails. In
numerical return models, realistic reproduction of historical data may
be achieved.

% In the domain of discrete-time returns' models,
% \gls{garch} and stochastic volatility models (SV) are the most popular, so in
% this thesis project we study a total of 7 intraday return series 
% using \gls{garch} and SV models and compare their performance in volatility
% forecast.
In addition to the model of a single asset, the correlation between
a group of assets is also of great interest. For example, in principle
component analysis, one wishes to identify a number of factors that
``drive'' the price evolution of a group of assets in the sense that
each of the assets' returns can be expressed as a linear combination
of the factors' returns. In this scenario, the eigenvalues of the
covariance matrix are the variances of the factors' returns and their
corresponding orthogonalized eigenvectors give the composition of the
factors, i.e. the coefficients with which the factors are contructed
as a linear combination of the assets.

Therefore, in this thesis we also study the elements and the
eigenvalues distribution, as well as the eigenvectors composition of
the covariance matrix. When the matrix is constructed from returns
with simple Gaussian distribution, the matrix is termed a Wishart matrix
and has been studied extensively in the literature. If the returns
have L\'evy distributions, the matrix is termed Wishart-L\'evy and has
been studied to some extent, particularly regarding its eigenvalue
distribution \cite{politi2010}.

However, it is understood that real stock/index returns are much more
complicated than a straight-forward Gaussian or L\'evy distribution can
describe --- instead, one needs structured models. For this reason, we
are particularly interested in a covariance matrix obtained for
realistic return models. The so called \gls{garch}(1,1) model is a realistic
return model proven to have regularly varying tails
\cite{mikosch2000}. So we study properties of eigenvalues and
eigenvectors of such covariance matrices. Moreover, we also study how
auto-correlations in the returns influence the aforementioned
properties. Such auto-correlations, known as second-order
auto-correlations decay exponentially but may still leave footprints
in the covariance matrix.

This report is organized as follows: Chapter \ref{chp:PriceModels}
reviews some of the most influential return model. Parameters of the
models are fitted to a few intraday return series and the predictive
power of the models is compared. A calculation of the
unconditional distribution functions of \gls{sv} models,
especially in the case where the residual of the log-volatility
and the innovation of the return are correlated normal variates, is
also presented. In chapter \ref{chp:Gaussian} we investigate
distributions of eigenvalues of the Wishart matrix, and study the
influence of auto-correlated returns. In chapter
\ref{chp:CrossCorrelationFat} distributions of elements and
eigenvalues of the covariance matrix of identically specified
\gls{garch}(1,1) series are studied. Finally, chapter \ref{chp:summary}
summarizes the results. Supplementary materials are provided in the
appendices.

@@

\chapter{Covariance Matrix of GARCH(1,1) Returns}
\label{chp:CrossCorrelationFat}
\gls{garch} models and particularly \gls{garch}(1,1) models are widely used to
model financial return series of various time scales, ranging from
daily and monthly returns that have been studied extensively in the
literature to intraday returns that we have selectively investigated
in chapter \ref{chp:PriceModels}. One particularly nice feature of
\gls{garch}(1,1) models is that they have regularly varying tails (power-law
tails) \cite{Mikosch2009, mikosch2000} even when the innovations
(denoted $z_t$ in the following text) are normally distributed ---
something not shared by other classes of models (e.g. not by
\gls{sv} models, as shown in section \ref{sec:XieCalc}) but well
documented for realistic returns data by empirical studies
\cite{Mantegna2000, Potters2003, Guhr2007}. However, what this tail
behavior implies for the covariance matrix is much less understood,
especially when the return series of the covariance matrix are
auto-correlated.

So in this chapter we consider the covariance matrix of N
identically specified, possibly auto-correlated \gls{garch}(1, 1)
processes:
\begin{eqnarray}
  r_{it} &=& \phi r_{i, t-1} + \epsilon_{it} \nonumber \\
  \epsilon_{it} &=& \sigma_{it} z_{it} \label{eq:garch_spec}
\end{eqnarray}
where $i=1,2,...,N$; $t=1,2,...,T$; $z_{it}$ is independent,
identically distributed, and
\begin{eqnarray*}
  \sigma_{it}^2 &=& \alpha_0 + \alpha_1 z_{i, t-1}^2 + \beta_1
  \sigma_{i,t-1}^2
\end{eqnarray*}
Mikosch and Starica showed in \cite{mikosch2000} that a
\gls{garch}(1,1) process satisfying
\begin{eqnarray*}
  \alpha_0 &>& 0 \\
  \E\ln(\alpha_1 Z^2 + \beta_1) &<& 0 \\
\end{eqnarray*}
and
\begin{eqnarray*}
  \E[(\alpha_1 Z^2 + \beta_1)^{p/2}] &\geq& 1 \\
  \E|Z|^p \ln|Z| &\leq& \infty
\end{eqnarray*}
for some $p > 0$, is stationary and has regularly varying tails. The
tail exponent $\alpha$ is determined by:
\begin{equation}\label{eq:garch_alpha}
  \E[(\alpha_1 Z^2 + \beta_1)^{\alpha/2}] = 1
\end{equation}
Here $Z$ is a random variable that has the same distribution as
$z_{it}$. In our simulations described hereafter, $z_{it}$ and $Z$
have standard Gaussian distribution. In this section and the next, we
first study situations where no auto-correlations are present among
the returns, i.e. $\phi = 0$; then in section \ref{sec:garch_nonzero_phi} we
look at how auto-correlations change the picture.

With regularly varying tails, the eigenvalue distribution of a
covariance matrix built from \gls{garch}(1,1) returns is expected to
differ from the Marcenko-Pastur law discussed in section
\ref{sec:wishart_eigen_dist}. In figure \ref{fig:garch_ev_cii} we
simulate N=50 independent \gls{garch}(1,1) 
returns series, each with identical parameters, namely $\alpha_0 =
2.3\times 10^{-6}$, $\alpha_1 = 0.15$, $\beta_1 = 0.84$, $\phi = 0$
and T=$8\times10^4$ time steps, then we build the covariance
matrix as
\begin{eqnarray*}
  \mtx C &=& {1 \over T^{2/\alpha}} \mtx{RR'}
\end{eqnarray*}
where $\mtx R$ is an $N\times T$ matrix, whose elements $r_{it}$ are
specified by equation \ref{eq:garch_spec} with $\phi = 0$. The
normalization factor $1 \over T^{2/\alpha}$ has been chosen such that the
eigenvalue distribution is independent of $T$ in the limit $T \to
\infty$ \cite{politi2010, Cizeau1994}.

The \gls{pdf} of the eigenvalue distribution of C is plotted in figure
\ref{fig:garch_ev_cii_linear} and \gls{cdf} of the distribution is
plotted on log-log scale in \ref{fig:garch_ev_cii_loglog}.
\begin{figure}[htb!]
  \centering
  \subfigure[]{
    \includegraphics[scale=0.40, clip=true, trim=104 271 111
    216]{../pics/garch_ev_cii.pdf}
    \label{fig:garch_ev_cii_linear}
  }
  \subfigure[]{
    \includegraphics[scale=0.40, clip=true, trim=104 274 104
    226]{../pics/garch_ev_cii_loglog.pdf}
    \label{fig:garch_ev_cii_loglog}
  }
  \caption{\small \it \ref{fig:garch_ev_cii_linear}: Eigenvalues and
    Diagonal elements' distribution of a covariance matrix
    built from independent GARCH return series. Blue: PDF of
    eigenvalues; Green: PDF of diagonal elements; Red: PDF of a
    $\alpha$-stable distribution fitted to the diagonal
    elements. \ref{fig:garch_ev_cii_loglog}: CDF of the same
    quantities in 10-based log-log scale.}
  \label{fig:garch_ev_cii}
\end{figure}
Also plotted in the same figure is the distribution of the diagonal
elements of $\mtx C$. It is clear from the figure that the two PDFs
coincide, implying $\mtx C$ is diagonal. This is further confirmed by
figure \ref{fig:garch_cij} which shows the distribution of the
non-diagonal elements of C. One can see the non-diagonal elements are
distributed symmetrically around 0 with a very small width in
comparison to the distribution of the diagonal elements --- in fact, 1
order of magnitude smaller ($2.14\times10^{-5}$ v.s. $7.31\times
10^{-4}$). Hence $\mtx C$ is very close to a diagonal matrix.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=95 253 91
    223]{../pics/garch_cij.pdf}
  \caption{\small \it Distribution of the non-diagonal elements of
    the covariance matrix. Blue: PDF of the non-diagonal
    elements; Green: $\alpha$-stable distribution fitted to the
    non-diagonal elements' PDF.}
  \label{fig:garch_cij}
\end{figure}

Figure \ref{fig:garch_ev_cii} also shows the two curves are well fitted
by an $\alpha$-stable distribution. An estimate of the L\'evy index
$\alpha$ of the stable distribution is also obtained via fitting,
$\alpha \approx 1.38$, as shown in figure \ref{fig:garch_ev_cii}. This
is really an expected result for the diagonal elements. Using
$\alpha_1 = 0.15$, $\beta_1 = 0.84$, which are the values used for
simulating the \gls{garch} returns, one can obtain, by solving equation
\ref{eq:garch_alpha}, $\alpha=2.96$. Then according to Mikosch and Starica
\cite{mikosch2000}
\begin{eqnarray*}
  P(|r_t| > x) &\sim& {\E(z^\alpha) c_0 \over x^\alpha} \\
  P(|r_t|^2 > x) &\sim& {\E(z^\alpha) c_0 \over x^{\alpha/2}} \\
\end{eqnarray*}
for some constant $c_0$. Now that $P(|r_t|^2 > x)$ has power-law tail
behavior with power $\alpha/2 < 2$, one can deduce
\begin{equation}
  \label{eq:stable_CLT}
  \begin{aligned}
    \sum_{t=1}^T r_t^2 &\xrightarrow{d} S(\alpha/2,
    1, \gamma, \mu) \text{ as $T \to \infty$}
  \end{aligned}
\end{equation}
where $\xrightarrow{d}$ denotes convergence in distribution, and
$S(\alpha/2, 1, \gamma, \mu)$ denotes an $\alpha$-stable distribution
with parameters $(\alpha/2, 1, \gamma, \mu)$. Here $\alpha/2$ is the
L\'evy index, 1 is the asymmetry, $\gamma$ is the scaling parameter
and $\mu$ is the mean value of the distribution. Asymmetry being 1
means a random variable so distributed only takes positive values
\cite{Bilik2008, Embrechts1997}.

The mean $\mu$ in equation \ref{eq:stable_CLT} is given by
\begin{eqnarray*}
  \mu &=& T \E(|r_t|^2) \\
  &=& T{\alpha_0 \over 1 - \alpha_1 - \beta_1}
\end{eqnarray*}
where we have used the result $\E(|r_t|^2) = \alpha_0 / (1 - \alpha_1
- \beta_1)$ \cite{Bollerslev86}. The scaling parameter $\gamma$ in
\ref{eq:stable_CLT} is determined by the limit \cite{Bilik2008}
\begin{eqnarray*}
  \lim_{T\to\infty} {T \E(z^{\alpha/2}) c_0 \over \gamma^{\alpha/2}}
  &=& C_{\alpha/2}
\end{eqnarray*}
where
\begin{eqnarray*}
  C_{\alpha/2} &=& \left( \int_0^\infty {\sin x \over x^{\alpha/2}} dx
  \right)^{-1} \\
  &\approx& {1 \over \sqrt{2 \pi}}
\end{eqnarray*}
Therefore
\begin{eqnarray*}
  \gamma^{\alpha/2} &=& \sqrt{2\pi} T \E\left(
    |z|^{\alpha/2}
  \right) c_0 \\
  \gamma &=& (2\pi)^{1/\alpha} T^{2/\alpha} \left(\E
    |z|^{\alpha/2}
  \right)^{2/\alpha} c_0^{2/\alpha}
\end{eqnarray*}
Here we note that an $\alpha$-stable distribution $S(\alpha, \beta,
\gamma, \mu)$ has characteristic function \cite{Guhr2007}
\begin{eqnarray*}
  \varphi(k; \alpha, \beta, \gamma, \mu) &=& \exp\left[
    i\mu k - \gamma^\alpha |k|^\alpha \left(
      1 - i \beta {k \over |k|} \tan{\pi \alpha \over 2}
    \right) \right] \text{ for $\alpha \neq 1$}
\end{eqnarray*}
from which we see $\varphi(ak; \alpha, \beta, \gamma, \mu) = \varphi(k;
\alpha, \beta, a\gamma, a\mu)$, implying that, if $x \sim S(\alpha,
\beta, \gamma, \mu)$, then $ax \sim S(\alpha, \beta, a\gamma, a\mu)$.

Now that
\begin{eqnarray*}
  \sum_{t=1}^T r_t^2 &\xrightarrow{d} S(\alpha/2,
  1, \gamma, \mu) \text{ as $T \to \infty$}  
\end{eqnarray*}
we have
\begin{eqnarray}
  C_{ii} &=& {1 \over T^{2/\alpha}}\sum_{t=1}^T r_{it}^2 \\
  &\xrightarrow{d}&
  S(\alpha/2, 1, \gamma_D, \mu_D) \text{ as $T \to \infty$}
  \label{eq:garch_cii_dist}
\end{eqnarray}
where
\begin{eqnarray}\label{eq:garch_wishart_cij_params}
  \gamma_D &=& (2\pi)^{1/\alpha} \E\left(|z|^{\alpha/2}
  \right)^{2/\alpha} c_0^{2/\alpha} \\
  \mu_D &=& {\alpha_0 \over 1 - \alpha_1 - \beta_1} T^{1 - 2/\alpha}
  \nonumber
\end{eqnarray}
So the diagonal elements of the covariance matrix converge to
an $\alpha$-stable distribution with L\'evy index $\alpha/2 \approx
1.48$. This is comparable to the index value 1.38 obtained by
fitting. Considering the slow convergene of regularly varying tails,
this is a reasonably good match.
